# Will You Make Money?

> No one can win at roulette unless he steals money from the table while the croupier isn’t looking.
>
> --- Albert Einstein (possibly)

The development of probability theory is historically linked to attempts to understand games of chance, especially ones in which money was involved (see for example, [here](http://sites.math.rutgers.edu/~cherlin/History/Papers2000/cheng.html). Sometimes betting money on an uncertain outcome falls under the name of gambling; other times it’s dignified with the name investment or "smart business decision." But regardless of the label, there are smarter and less smart ways to play money games. 

## Battle of the Bills {-}

Let’s recall a distinction we made earlier in this course about deterministic and stochastic or random processes. This time, we’ll think about two different bets you make with your friend. In the first bet, you and your friend are debating whether it was Bill Paxton or Bill Pullman in the movie Apollo 13. To make the game interesting, you bet two dollars. You look it up on the internet, and find that it was indeed Paxton. One of you wins. Do you feel the need to check again? Probably not. This particular question, although you may not have known the answer for sure, has only one possible answer.

Now consider another bet, this time for three bucks! You and your friend are walking down the street debating the “merits” of mint chocolate chip vs. cookies and cream as ice cream flavors. You claim that mint chocolate chip is the more popular flavor, and decide to ask the first passer-by which flavor they think is better. Suppose they don’t just ignore you, thinking you’re a nutcase, and they answer cookies and cream. Are you satisfied with this one answer? Or do you feel the need to ask another pedestrian? And how many?

We might say that the variable “BPA”, which stands for “which Bill P. starred in Apollo 13?” has a deterministic answer, but the variable “MCCoCAC,” which stands for “mint chocolate chip > cookies and cream?” can take on one of two answers (no ties allowed) depending on whom we ask. Because it is a random or stochastic variable, we have to talk about it using different terms. We might say something like, what proportion of people (in this neighborhood, say) prefer cookies and cream? Or what are the chances that the first person we ask will express that particular preference.

This may all sound like silly bets that are really just games between friends. But people make small and large money bets all the time, in everything from business and life decisions, to recreational games. In this chapter, we explore probability calculations that inform things like advertising, airplane booking, the job market, and march madness.


```{r, echo=FALSE}

cure_prob <- c(0.7,0.3)
approved_prob <- c(0.4, 0.6)
market_prob <- c(0.1, 0.9)


wheelan_cancer <- data.frame(
   Cure = c(rep("No",700), rep("Yes",300)),
   Approved = c(rep(NA,700), rep("No",120), rep("Yes",180)),
   First2Market = c(rep(NA,820), rep("No", 18), rep("Yes",162)),
   Payout = c(rep(250000,700), rep(0,120), rep(0, 18), rep(25000000,162))
)

wheelan_tree <- rpart(
  Payout ~ ., 
  data = wheelan_cancer, 
  method = "anova"
)
```

```{r, echo=FALSE}
rpart.plot(wheelan_tree, type = 5)
```

Consider the diagram on page 83 of the Naked Statistics chapter that you read at the beginning of this module. This diagram shows the drug approval process from investment to pay off. At each node of the graph, there are two possibilities: first, you may or may not develop a drug that cures a particular disease. Next, even if the drug works, it may or may not get approved. And even if it gets approved, it may or may not make it to the market. At each branch, there are two possibilities with estimated probabilities (note: how do we estimate these probabilities in practice?). All of the branches lead to 5 possible outcomes, with different pay-offs. Using what we know about all of the possibilities, all of their probabilities, and the potential payoffs, we can estimate what is called “expected value”. [walk through expected value calculation]. This value represents the average amount of money we would expect to make, if we invested in a lot of drugs, that all went through this same process independently. 

Question to think about: even though the expected average pay-off is $4,225,000, which is more than 4 times the original investment, would you want to make this investment for any single drug? Why or why not?

Now that we’ve made this calculation empirically, it might be helpful to simulate it in R! https://a3sr.shinyapps.io/Drug 

Week 9 Video 1
Let’s now return to the example that you read about last week (Wheelan, chapter 5):
In 1981, Schlitz brewing company, now defunct but at one time the largest beer producer in the US, ran a bold advertising campaign. During the Super Bowl, Schlitz ran a live blind taste test against one of its competitors, Michelob. 100 Michelob drinkers participated in the taste test, which aired LIVE. The advertisement slot itself cost a lot of money. Schlitz could have just run a funny ad involving puppies on the beach, so why take a risk with a taste test that could conceivably have gone badly. How could Schlitz have been so confident that their beer would be preferred? 

THINK ABOUT IT QUESTION: What information would you need to know to advise the Schlitz brewing company about running such an ad? (please take a few minutes before continuing on, to try to list this information on your own).

As discussed in the book, some things we would need to know are:
Intended sample size for taste test
Actual proportion of Michelob drinkers who would prefer Schlitz in a blind taste test
Acceptable outcome of live taste test for promoting Schlitz beer
Rules of mathematical probability

Run and discuss simulation in Shiny: https://a3sr.shinyapps.io/Schlitz/
Week 10 Reading 1
In the last two weeks, we’ve explored the concept of expected value. As you may have noticed, expected value can be very useful for making decisions, but it does not tell the whole story. In the college majors example, expected career earnings for each possible major were only useful in describing what happens on average, for a lot of people going through the job market. However, we still had a lot of uncertainty about what would happen to any single individual who pursued a career in acting or accounting. Similarly, we would not have been as confident running the Schlitz commercial with a sample of 10.

This brings us to one of the most important concepts in probability and statistics: the “law of large numbers”. Random variables are by definition, random, so we cannot predict any single outcome -- we can only say how often we expect a particular outcome to occur if we observe a lot of data. Thus, you might see how expected value is particularly useful in situations where the same process gets repeated many times. In the airplane overbooking example, there may be a reasonable probability that airlines will lose money on any given flight. But airlines can make probability calculations to ensure that they earn money than they lose overall. 

The usefulness of the law of large numbers is particularly evident in gambling. Many people go to casinos and play the same games over and over again, which each (theoretically) have consistent probabilities of winning or losing. One person may win and another may lose, but if a lot of people play a game of roulette (for example), casinos can estimate approximately what proportion will win and how many will lose. And, the more games that are played, the more confident the casino can be about their predicted proportion of wins/losses.

As you’ve probably heard, casinos are designed such that the house always has the advantage. In other words, every casino game is designed such that the casino always has a slightly better chance at winning than the player. That doesn’t mean that players can’t win individual games, it just means that, assuming enough people play the game, the casino will almost surely win more games than all the individual players. 

This construct tempts us to believe that we can cheat the system. There is a long and interesting history of people trying to figure out betting strategies that will give them back the advantage. The Martingale betting strategy is one example. Please read the following article to learn more about it! https://www.roulettesites.org/strategies/martingale/ 

Week 10 Reading 2
As a final example of how we can use probabilistic thinking to inform betting decisions, let’s consider March Madness. March Madness is a basketball tournament, where 64 teams compete. In the first round, 32 games are played. The winners of those games then play each other in the 16 second round games. This continues until a single team is named victorious. Every year, people make bets on which teams they think will win each match-up. Data from regular season games can be used to estimate the probability of any team winning a particular match-up, and this information is readily available to fans. Yet, despite the wealth of data that is available to make game predictions, no one has ever correctly predicted all of the sequential game winners in the tournament. In fact, only one person has ever predicted the first two rounds perfectly (this happened in 2019!). How is this possible? In order to learn more about the complexities of choosing a March Madness bracket, please read the following article: https://www.scientificamerican.com/article/how-much-math-do-you-need-to-win-your-march-madness-pool/ 

Week 10 Video 1
Now that you know a little more about March Madness, let’s just consider the probability of predicting the entire first round correctly. To start, we will (incorrectly) assume that every team has a 50% chance of winning their first round game. In this case, we have a 50% chance of guessing the correct winning team. Given that there are 32 games in the first round, how could we calculate the probability of predicting them all correctly? [walk through calculation, drawing a tree diagram, similar to those in the first two weeks of the module]
